# 02. Preprocessing

## Setup

```{r}
# Load libraries
library(tidyverse)

# Load the data
data <- read.csv("../data/raw/train.csv")

# View data
head(data)
dim(data)
str(data)
```


## Column elimination
According to the conclusions of the Exploratory Analysis, we will eliminate the following columns:
- `building_id`
- `plan_configuration`
- `legal_ownership_status`
- `count_families`
- `has_secondary_use_use_police`
- `has_secondary_use_gov_office`
- `has_secondary_use_health_post`
- `has_secondary_use_industry`
- `has_secondary_use_institution`
- `has_secondary_use_school`
- `has_secondary_use_other`

```{r}
data <- data %>%
  select(-building_id, -plan_configuration, -legal_ownership_status, -count_families, -has_secondary_use_use_police, -has_secondary_use_gov_office, -has_secondary_use_health_post, -has_secondary_use_industry, -has_secondary_use_institution, -has_secondary_use_school, -has_secondary_use_other)

# View data
head(data)
dim(data)
str(data)
```

As we can see, after eliminating the mentioned columns, the data has 26 columns.

## Label encoding

## Column unification

## Principal Component Analysis

## Imbalanced data handling