# Exploratory Analysis

To conduct an exploratory analysis of the data we need to do several things
- Load the data
- Check the data types
- Check for missing values
- Check for duplicates
- Check for data distribution
- Check for data correlation with the target variable to clean data
- Visualize and interpret the data to understand it better

## Load the data

```{r}
# Load libraries
library(tidyverse)
# Load the data
data <- read.csv("../data/raw/train.csv")
# View data
head(data)
dim(data)
```

We can already see some columns that will need some processing and analysis to understand its content. We can identify so far:
- Percentages columns in integers
- Categorical columns as strings (chr)
- Boolean columns as integers, which need duplication checking

We will take this into account later for data visualization and interpretation


## Check duplicate values

```{r}
# Check for duplicates
duplicated_rows <- data[duplicated(data),]
duplicated_rows

duplicated_columns <- data[,duplicated(t(data))]
duplicated_columns
```

Fortunately the data seems to be quite clean and does not contain any duplicates, or exact duplicates. But some columns such as has_superstructure_rc_non_engineered and has_superstructure_rc_engineered seemed to be opposite in description and presented the same first few values, so we will need to look further into them later on.


## Check Null values

```{r}
# Check for missing values
missing_values <- data %>% summarise_all(~sum(is.na(.)))
missing_values
```

Same as previously, data seems to be quite clean in this regard with no missing values.


## Data correlation


## Data visualization and interpretation

Now we will try to extract some meaning behind the data, to understand it better before we proceed with the processing. This is due to potential significant information being hidden in the data that we can use to improve our models down the line.

### Percentages

```{r}
# Check value limits for percentages colums (contain "percentage" in colname)
percentages_columns <- colnames(data)[grepl("percentage", colnames(data))]
summary(data[, percentages_columns])
```

As we can see the 3rd quartile and max values are pretty far apart, so we can use a boxplot to check for outliers.

```{r}
# Boxplot for percentages columns
par(mfrow=c(1,2))
boxplot(data$area_percentage, main="Area Percentage", col="lightblue")
boxplot(data$height_percentage, main="Height Percentage", col="orange")
```
We expected to extract significant information from these plots, imagining it would indicate the area of the building damaged and a similar meaning for height. However seeing that the 3rd quartile are low for both and maximum values are not that much higher either, we see no descriptive conclusion for this data.

### Categorical

```{r}
# Check list of values for categorical columns (<chr>)
categorical_columns <- colnames(data)[sapply(data, is.character)]
```

```{r}
# We check the different values of each column by showing unique values.Represent by showing bar plots of each of the columns. x axis being values, y axis being counts. Add colors per column and add a number a top of each bar.
for (col in categorical_columns) {
  barplot(table(data[, col]), col=rainbow(length(table(data[, col]))), main=col)
}
```

After observing the position variable, we wonder what could it mean and if it is related to distance to the epicenter. To check this we can group by position and check the mean of the target.

```{r}
# Plot position variable against target variable.
position_target <- data %>% group_by(position) %>% summarise(mean_target = mean(damage_grade))
ggplot(position_target, aes(x=position, y=mean_target)) + geom_bar(stat="identity")
```

Unfortunately there does not seem to be a clear relationship between position and damage_grade.

We also want to check what could foundation types mean and wether some are "stronger" or better than others which could be useful for preprocessing data.

```{r}
# Plot foundation_type variable against target variable.
foundation_target <- data %>% group_by(foundation_type) %>% summarise(n=n(), mean_target = mean(damage_grade))
foundation_target
```

```{r}
# barplot for each foundation type
ggplot(foundation_target, aes(x=foundation_type, y=mean_target)) + geom_bar(stat="identity")
```

We believe foundation type is closely related to damage grade, not only because of intuition but because the mean damage grade is different for each foundation type. We will keep this in mind for the next stages.

#### Other categorical columns

As far as other categorical columns go, `plan_configuration` and `legal_ownership_status` are very clear to us that they will only add noise or non-valuable information to further analysis as a very high percentage (close to 100%) are of the same value and we believe will only increase compute time when building predictive models.

As for the other columns, we will keep them for now but do not see any clear correlation with the target variable just yet from intuition only.

### Boolean


## Analysis with specialized utilities


## Conclusions